---
title: "hw04"
author: "Prerna Gupta"
output: 
  html_document:
  always_allow_html: yes
  keep_md: true
---

```{r allow errors, echo = FALSE}
knitr::opts_chunk$set(error = TRUE)
```


```{r, echo = FALSE, warning = FALSE, message = FALSE}
library(tidyverse)
guest <- read_csv("https://raw.githubusercontent.com/STAT545-UBC/Classroom/master/data/wedding/attend.csv")
email <- read_csv("https://raw.githubusercontent.com/STAT545-UBC/Classroom/master/data/wedding/emails.csv")
library(singer)
library(ggplot2)
library(gapminder)
knitr::opts_chunk$set(fig.width=4, fig.height=3, warning = FALSE, fig.align = "center")
```
.

# Exercise 1: Univariate Option 1

1. Make a tibble with one row per year, and columns for life expectancy for two or more countries.

```{r}
(gapminder_IndoChina <- gapminder %>%
                filter(country %in% c("China", "India")) %>%
                select(country, year, lifeExp))
```

```{r}
(gapminder_wide <- gapminder_IndoChina %>%
        pivot_wider(id_cols = c(-country, - lifeExp),
                    names_from = country,
                    values_from = lifeExp))
```


2. Take advantage of this new data shape to scatterplot life expectancy for one country against that of another.

```{r}
China_lifeEXp <- gapminder_wide %>%
                select(year, China)

India_lifeEXp <- gapminder_wide %>%
                select(year, India)

ggplot() +
  geom_point(data = China_lifeEXp, aes(year, China, colour = "China_lifeEXp")) +
  geom_point(data = India_lifeEXp, aes(year, India, colour = "India_lifeEXp")) +
  xlab("Year") + ylab("Life Expectancy") + ggtitle("Life Expectancy of China and India") +
  scale_colour_manual("country", values = c("China_lifeEXp"= "red", "India_lifeEXp" = "blue")) +
  theme_bw()

```


3. Re-lengthen the data.

```{r}
gapminder_wide %>%
  pivot_longer(cols = -year,
               names_to = "country",
               values_to = "lifeExp")
```


# Exercise 2 : Multivariate Option 1

1. Make a tibble with one row per year, and columns for life expectancy and GDP per capita (or two other numeric variables) for two or more countries.

```{r}
(gapminder2_wider <- gapminder %>%
  filter(country %in% c("China", "India")) %>%
  select(year, country, lifeExp, gdpPercap)  %>%
  pivot_wider(id_cols = year,
              names_from = country,
              names_sep = "_",
              values_from = c(lifeExp, gdpPercap)))

```

2. Re-lengthen the data.

```{r}
gapminder2_wider %>%
  pivot_longer(cols = -year,
               names_to = c(".value", "country"),
               names_sep = "_")
```


Exercise 3: Table Joins

3.1 : For each guest in the guestlist (guest tibble), add a column for email address, which can be found in the email tibble.
```{r}
(email_name <- email %>%
          separate_rows(guest, sep = ", ") %>%
          rename(name = guest))
  
```

```{r}
guest %>%
  left_join(email_name, by = "name")
```

3.2 : Who do we have emails for, yet are not on the guestlist?

```{r}
(guest_email <- email_name %>%
                anti_join(guest, by = "name") %>%
                select(email, name))

```
Three guests are on the email list but not on the guest list


3.3 : Make a guestlist that includes everyone we have emails for (in addition to those on the original guestlist)
```{r}
(all <- full_join(guest, email_name, by = "name"))
```


