---
title: "hw_02"
output: 
  github_document:
    keep_md : true
editor options: 
  chunk_output_type: inline 
---

```{r load packages, warning = FALSE, message = FALSE}
library(gapminder)
library(tidyverse)
```

Exercise 1: Basic dplyr


```{r}
# 1.1 Using filter() to subset gapminder data to three countries of choice in the 1970’s.
gapminder %>%
filter (year > 1970 & year < 1980 | year == 1970,  
           country == "China" | country == "India" | country == "Pakistan")
```

```{r}
# 1.2 Using the pipe operator %>% to select “country” and “gdpPercap” from filtered dataset in 1.1. 
gapminder %>%
filter (year > 1970 & year < 1980 | year == 1970,  
           country == "China" | country == "India" | country == "Pakistan") %>%
 select(country, gdpPercap)

```

```{r}
# 1.3 Filtering gapminder to all entries that have experienced a drop in life expectancy.
gapminder %>%
 group_by(country) %>% 
 mutate(lifeExp_decrease = lifeExp - lag(lifeExp, 1, order_by = year)) %>%
 drop_na() %>%
 filter(lifeExp_decrease < 0)
```

```{r}
# 1.4 Filtering gapminder so that it shows the max GDP per capita experienced by each country.

gapminder %>%
  group_by(country, continent) %>%
  summarise(maxGDPperCap = max(gdpPercap))
```

```{r}
# 1.5 Producing a scatterplot of Canada’s life expectancy vs. GDP per capita using ggplot2, without defining a new variable.

gapminder %>% 
  filter(country == "Canada") %>% 
  ggplot(aes(gdpPercap, lifeExp)) + 
  geom_point() +
  scale_x_log10("GDP per capita", labels = scales::dollar_format()) + 
  labs(title="Canada", y="Life Expectancy")
  
```


Exercise 2: Explore individual variables with dplyr

I will be exploring two variables - life expectancy and continents

To understand the possible values in the variable "continents" I will explore the number of observation per continent and how many unique countries are there in each continent.

```{r}
gapminder %>%
  group_by(continent) %>%
  summarise(number_of_observations = n(),
            number_of_countries = n_distinct(country))
``` 
Asia and Africa have the largest number of countries.

Now I will explore the variable life expectancy. I will first explore the range of the variable by continent in year 2007
```{r}
gapminder %>%
  filter(year == "2007") %>%
  group_by(continent) %>%
  summarise(min_lifeExp = min(lifeExp), max_lifeExp = max(lifeExp)) %>%
  mutate(range = max_lifeExp - min_lifeExp)
```
The range is largest for Asia and Africa. One of the reasons for this can be large number of countries as we saw in the previous table. However, Europe has only 3 countries less than Asia but its range of life expectancy is much smaller than Asia. 

Now I will explore the summaries of lifeExp by continent in the year 1952 and 2007 to find typical values through median and mean and spread via standard deviation 
```{r}
gapminder %>%
  filter(year %in% c(1952,2007)) %>%
  group_by(continent, year) %>%
  summarize(median_lifeExp = median(lifeExp),
            mean_lifeExp = mean(lifeExp, na.rm = TRUE),
            standard_deviation_lifeExp = sd(lifeExp, na.rm = TRUE))
```

Now I will use boxplot for a visual exploration of the typical values and the spread over the years

```{r}
gapminder %>% 
  select(lifeExp, year) %>%
  mutate(year= factor(year)) %>% 
  ggplot(aes(year, lifeExp)) +
  scale_y_log10("Life Expectancy") +
  geom_boxplot() + xlab("Year")
```

Now I will explore the distribution of life expectancy in one country ie. India over the years using bar graphs. 

```{r}
gapminder%>%
  filter(country == "India")%>%
  select(lifeExp, year) %>%
  mutate(year = factor(year)) %>%
  ggplot(aes(year, lifeExp)) +
  scale_y_continuous ("Life Expectancy") +
  geom_bar(stat = "identity") + xlab("Year")
```

Exercise 3: Explore various plot types. 

```{r}
Orange %>% 
  group_by(Tree) %>% 
  select(age, circumference) %>% 
  ggplot(aes(age, circumference, color=Tree)) +
  geom_line() +
  xlab("age") + 
  labs(title="Age by Circumference of Orange Trees", y="Circumference") +
  theme(plot.title = element_text(hjust = 0.5))
```
This graph shows how the circumference of trees increase as their age increases. 

```{r}
Orange %>% 
  filter(Tree == "2") %>% 
  select(age, circumference) %>% 
  ggplot(aes(circumference, age)) +
  geom_bar(stat = "identity") + 
  labs(title="Circumference by Age of Orange Tree 2") +
  theme(plot.title = element_text(hjust = 0.5))
```

Bonus Exercise
```{r}
filter(gapminder, country == c("Rwanda", "Afghanistan"))
```
"==" is a logical operator that has here converted country into a vector of length 2 consisting of Rwanda and Afghanistan. "==" is meant to compare if two things are exactly equal. Since, here the vectors are not of equal length, so Vectors- Rwanda and Afghanistan are recycled.  

To fix this we can use %in% which checks whether or not the object (here country) is contained in the other object(c(Rwanda, Afghanistan) or not and then filter drops the NA values.So we get the following table.
```{r}
gapminder %>%
    filter (country %in% c("Rwanda", "Afghanistan"))
```

